---
title: "Tutorial 4"
author: "Jack Halverson"
date: "9/27/2021"
output: html_document
---

```{r}
options(java.parameters = "-Xmx2G")

library(r5r)
library(osmextract)
library(tidyverse)
library(sf)
library(ggthemes)
library(ggspatial)
library(tigris)
library(wesanderson)
library(tidytransit)
```

### Load Networks
```{r}
dir.create("networks")

download.file("https://app.mecatran.com/urb/ws/feed/c2l0ZT1zbG90cmFuc2l0O2NsaWVudD1zZWxmO2V4cGlyZT07dHlwZT1ndGZzO2tleT0zZTMwMzM1OTRiMTE2NzA0N2IxNjQwNjA0ZjQwMGMzMzdiM2E1MTQ0", file.path("networks","SLOgtfs.zip"), mode = "wb", quiet=TRUE)
```

### Street Network Data
```{r}
SLO_file <- oe_match("San Luis Obispo")
```

### Download data to networks folder and read layer of lines representing the street network
```{r}
SLO_streets <- oe_read(SLO_file$url, 
                   provider = "openstreetmap_fr", 
                   download_directory = "networks", 
                   layer = "lines", 
                   quiet = TRUE) %>%
  filter(!is.na(highway))
```

### Test Plot
```{r}
ggplot(SLO_streets) +
  geom_sf()
```

### Mask Streets to be only in the City Limits and Project onto Coordinate System
```{r}
CA5_state_plane <- "+proj=lcc +lat_1=34.03333333333333 +lat_2=35.46666666666667 +lat_0=33.5 +lon_0=-118 +x_0=2000000 +y_0=500000 +ellps=GRS80 +units=m +no_defs"

SLO_city_limits <- places("California") %>%
  filter(NAME == "San Luis Obispo") %>%
  st_transform(crs = st_crs(SLO_streets))


SLO_streets <- SLO_streets[SLO_city_limits,]

ggplot(SLO_streets) +
  geom_sf() +
  coord_sf(crs = CA5_state_plane)
```

### Load School Locations and rename osm_id column to be "id" for the r5r package to read correctly
```{r}
SLO_schools <- oe_read(SLO_file$url, 
                   provider = "openstreetmap_fr", 
                   download_directory = "networks", 
                   layer = "points", 
                   quiet = TRUE) %>%
  filter(str_detect(other_tags, '"amenity"=>"school"')) %>%
  st_filter(SLO_city_limits) %>%
  rename(id = osm_id)
```

### Create grid of points (hexagons) across entire city to compare travel times to/from
```{r}
grid <- st_sf(st_make_grid(SLO_city_limits, 
                           square = FALSE, 
                           n = c(100,100),
                           what = "polygons")) %>%
  st_filter(SLO_city_limits) 

colnames(grid) <- "geometry"
st_geometry(grid) <- "geometry"

grid <- grid %>%
  mutate(id = seq(1, length(grid$geometry), by=1))

grid_points <- st_centroid(grid)

ggplot(grid) +
  geom_sf() +
  geom_sf(data = grid_points, alpha = 0.5) +
  geom_sf(data = SLO_streets) +
  coord_sf(crs = CA5_state_plane) + 
  theme_map()
```

### Set up r5r core
```{r}
r5r_core <- setup_r5("networks", verbose = FALSE)
```

### Calculate travel time matrix
```{r}
ttm <- travel_time_matrix(r5r_core = r5r_core,
                          origins = SLO_schools,
                          destinations = grid_points,
                          mode = c("WALK", "TRANSIT"),
                          departure_datetime = as.POSIXct("15-11-2021 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S"),
                          max_walk_dist = 1000,
                          max_trip_duration = 480,
                          verbose = FALSE)
```

### Reformat ttm to have each row be an origin and columns for the schools
```{r}
tt_wide <- ttm %>%
  pivot_wider(names_from = fromId, 
              names_prefix = "from", values_from = travel_time) %>%
  rename(id = toId) %>% 
  merge(grid) %>%
  replace(is.na(.), 999) %>%
  rowwise() %>%
  mutate(from_any = min(c_across(starts_with("from")), na.rm = TRUE))

st_geometry(tt_wide) <- "geometry"
```

### Plot travel time to nearest school
```{r}
ggplot(SLO_streets) +
  geom_sf(data = tt_wide, 
          aes(fill = from_any), 
          color = NA) +
  geom_sf(alpha = 0.5) +
  scale_fill_gradient2(low = "green", mid = "yellow", high = "red", 
                       midpoint = 30,
        name = "Transit Travel\ntime to the\nnearest school\n(minutes)") +
  coord_sf(crs = CA5_state_plane) +
  theme_map()
```

### Create Isochrones
```{r}
iso_pallete <- wes_palette("Zissou1", n = 5)

iso10min <- tt_wide[tt_wide$from_any < 11,] %>%
  st_union()

iso20min <- tt_wide[tt_wide$from_any < 21,] %>%
  st_union()

iso30min <- tt_wide[tt_wide$from_any < 31,] %>%
  st_union()

ggplot(SLO_streets) +
  geom_sf(data = iso30min, 
          aes(fill = "Area within 30 minutes"), 
          color = NA) +
  geom_sf(data = iso20min, 
          aes(fill = "Area within 20 minutes"), 
          color = NA) +
  geom_sf(data = iso10min, 
          aes(fill = "Area within 10 minutes"), 
          color = NA) +
  geom_sf(alpha = 0.5) +
  scale_fill_manual(values = c(iso_pallete[1], 
                               iso_pallete[3],
                               iso_pallete[5]),
        name = "Transit Travel\ntime to the\nnearest school\n(minutes)") +
  coord_sf(crs = CA5_state_plane) +
  theme_map()
```

### Load transit stop locations
```{r}
SLO_transit <- read_gtfs(file.path("networks", "SLOgtfs.zip"))

transit_stops <- st_as_sf(SLO_transit$stops, 
                          coords = c("stop_lon", "stop_lat"), 
                          crs =st_crs(grid))
```

### Set up grid of points with stops data
```{r}
transit_grid <- grid %>%
  mutate(num_stops = lengths(st_covers(grid, transit_stops)))

transit_points <- st_centroid(transit_grid)

ggplot(transit_points) +
  geom_sf(aes(color = as.character(num_stops))) +
  scale_color_manual(values = c("gray", "cornsilk", "lightgreen"), 
                    name = "Number of\ntransit stops") +
  theme_void()
```

### Calculate accessibility - 10 minute walk to transit stop
```{r}
transit_access <- accessibility(r5r_core,
                        origins = transit_points,
                        destinations = transit_points,
                        mode = "WALK",
                        opportunities_colname = "num_stops",
                        decay_function = "step",
                        cutoffs = 11,
                        departure_datetime = as.POSIXct("15-11-2021 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S"),
                        max_walk_dist = 500,
                        time_window = 60,
                        percentiles = 50,
                        verbose = FALSE) %>%
  mutate(id = as.numeric(from_id)) %>%
  merge(grid)

st_geometry(transit_access) <- "geometry"

ggplot(transit_access) +
  geom_sf(aes(fill = accessibility), color = NA) +
  scale_fill_viridis_c(name = "Transit stops\nwithin 10-minutes\nwalk") +
  coord_sf(crs = CA5_state_plane) +
  theme_void()
```

### Calculate access with a decay function (more continuous) with "accessibility" decreasing by one half every five minutes
```{r}
transit_access2 <- accessibility(r5r_core,
                        origins = transit_points,
                        destinations = transit_points,
                        mode = "WALK",
                        opportunities_colname = "num_stops",
                        decay_function = "exponential",
                        cutoffs = 5,
                        departure_datetime = as.POSIXct("15-11-2021 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S"),
                        max_walk_dist = 500,
                        time_window = 60,
                        percentiles = 50,
                        verbose = FALSE) %>%
  mutate(id = as.numeric(from_id)) %>%
  merge(grid)

st_geometry(transit_access2) <- "geometry"

ggplot(transit_access2) +
  geom_sf(aes(fill = accessibility), color = NA) +
  scale_fill_viridis_c(name = "Accessiblity score") +
  coord_sf(crs = CA5_state_plane) +
  theme_void()
```

### Turn off r5r core to save memory
```{r}
stop_r5(r5r_core)
rJava::.jgc(R.gc = TRUE)
```

### Save grid of accessibility values to use in next assignment
```{r}
st_write(transit_access2, 'SLO_access.geojson', append=FALSE, quiet=TRUE )
```